---
title: "Progress Check-In #1"
name: "Thomas Walsh"
date: "March 11, 2026"
format: docx
---

Progress check 01 for the final project for BIFX-546: Machine Learning for Bioinformatics

General Outline:
1. Loading the Dataset
2. Initial cleaning
3. Visualizations
4. Summary statistic
5. Short progress note 

# Setting up datasets library and token usage
    Setting up token for more efficient load times

```{r}
#| eval: false

# install datasets library
pip install datasets

# use secret key from colab to authenticate Hugging Face login
from google.colab import userdata
userdata.get('HF_TOKEN')

```

# Loading the Dataset
    From Huggingface: https://huggingface.co/datasets/longevity-db/mouse-glioblastoma-snRNAseq 
    Without colab pro, I had to stream the data to avoid runtime/memory timeouts.

```{r}
#| eval: false

# Load the dataset
from datasets import load_dataset
import pandas as pd

dataset_name = "longevity-db/mouse-glioblastoma-snRNAseq"

try:
    # Load the dataset in streaming mode
    # This allows processing data without loading the entire dataset into memory
    # which is ideal for getting a 'smaller, more manageable portion'.
    streamed_dataset = load_dataset(dataset_name, streaming=True)

    # Assuming there's a 'train' split, we'll take the first 10000 examples.
    # You can adjust this number as needed.
    sample_size = 10000
    sampled_data = []
    for i, example in enumerate(streamed_dataset['train']):
        if i >= sample_size:
            break
        sampled_data.append(example)

    # Convert the sampled data to a Pandas DataFrame for easier manipulation
    df_sampled = pd.DataFrame(sampled_data)

    print(f"Successfully loaded a sample of {len(df_sampled)} examples.")
    print("\nFirst 5 rows of the sampled data:\n")
    display(df_sampled.head())

except Exception as e:
    print(f"Error loading dataset: {e}")
    print("Please ensure the dataset ID is correct, you have network access, and the HF_TOKEN (if required) is set in Colab secrets.")
```

# Initial cleaning:
    Check for missing values

```{r}
#| eval: false

print("Checking for missing values in df_sampled per column:")

found_missing_values = False
for col in df_sampled.columns:
  null_count = df_sampled[col].isnull().sum()
  if null_count > 0:
    print(f"Column '{col}': {null_count} missing values")
    found_missing_values = True

if not found_missing_values:
  print("No missing values found in df_sampled.")
```

# Initial cleaning:
    Check for unique values

```{r}
#| eval: false

import pandas as pd

single_values_data = {}
cols_to_drop = []

for col in df_sampled.columns:
    unique_vals = df_sampled[col].unique()
    if len(unique_vals) > 1:
        print(f"Column '{col}': There's more than one unique value.")
    else:
        single_values_data[col] = unique_vals[0] # Store the single unique value
        cols_to_drop.append(col) # Mark column for removal
        print(f"Column '{col}': Unique value is '{unique_vals[0]}'")

# Create a new DataFrame for columns with single unique values
df_single_unique_values = pd.DataFrame.from_dict(single_values_data, orient='index', columns=['Unique Value'])
print("\nDataFrame with single unique values:")
display(df_single_unique_values)

# Remove these columns from df_sampled
df_sampled = df_sampled.drop(columns=cols_to_drop)
print("\nOriginal df_sampled after removing columns with single unique values:")
display(df_sampled.head())

```

This creates a new dataframe with all the unique values for reference, but they're not necessary for data analysis and removes the unecessary value columns from the original dataset

# Visualizations
    Scatter plots

This shows a basic scatter plot depicting the number of Reads and the number of Genes overall

```{r}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

# Generate basic scatter plot
plt.figure(figsize=(10, 7))
sns.scatterplot(x='nCount_RNA', y='nFeature_RNA', data=df_sampled, alpha=0.6)

# plot labels
plt.title('Number of Reads vs. Number of Genes')
plt.xlabel('Reads (nCount_RNA)')
plt.ylabel('Genes (nFeature_RNA)')

plt.grid(True, linestyle='--', alpha=0.7)

# display plot
plt.show()
```

# Visualizations
    Scatter plots for each cell type

```{r}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

# Generate scatter plot
plt.figure(figsize=(12, 8)) # Increased figure size for better readability with more colors
sns.scatterplot(x='nCount_RNA', y='nFeature_RNA', hue='sub_celltype', data=df_sampled, alpha=0.6)

# Title and labels
plt.title('Number of Reads (nCount_RNA) and Number of Genes (nFeature_RNA) by Cell Type')
plt.xlabel('Reads (nCount_RNA)')
plt.ylabel('Genes (nFeature_RNA)')

# Add a grid for better readability
plt.grid(True, linestyle='--', alpha=0.7)

# Move the legend outside the plot area to prevent overlapping with data points
plt.legend(bbox_to_anchor=(1.05, 1), loc='upper left', borderaxespad=0.)

plt.tight_layout() # Adjust layout to prevent labels/legend from being cut off

# Display the plot
plt.show()
```

View the number of Female and Male observations:

```{r}
#| eval: false

print(df_sampled['sex'].value_counts())
```

First filter by sex
    Then generate the same scatter plot, but categorizing by sex
    This will help to visualize the differnece between Female or Male observations:

```{r}
#| eval: false

df_female = df_sampled[df_sampled['sex'] == 'female']
df_male = df_sampled[df_sampled['sex'] == 'male']

print(f"df_female created with {len(df_female)} observations.")
print(f"df_male created with {len(df_male)} observations.")
```

Next, I want to see the distribution of the number of Reads and Genes for each Cell Type
For this, I can use a violin plot:

```{r}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

# Figure and axes for the plot with an appropriate size
plt.figure(figsize=(16, 8)) # Increased figure size to accommodate many cell types

# Generate violin plot
sns.violinplot(x='sub_celltype', y='nCount_RNA', data=df_sampled, inner='quartile')

# Title and labels
plt.title('Distribution of Number of Reads (nCount_RNA) by Cell Type')
plt.xlabel('Cell Type')
plt.ylabel('Number of Reads (nCount_RNA)')

# Rotate the x-axis labels if they are overlapping to improve readability
plt.xticks(rotation=90)

# Adjust the plot layout to prevent elements from being cut off
plt.tight_layout()

# Display the plot
plt.show()
```

This helps to visualize how the nCount_RNA (number of reads) and nFeature_RNA (number of genes) differ across different cell types. The same type of violin plot can be used for the number of genes to see the distribution of those genes across all cell types:

```{r}
#| eval: false
import matplotlib.pyplot as plt
import seaborn as sns

# Create a figure and axes for the plot with an appropriate size
plt.figure(figsize=(16, 8)) # Increased figure size to accommodate many cell types

# Generate a violin plot
sns.violinplot(x='sub_celltype', y='nFeature_RNA', data=df_sampled, inner='quartile')

# Add a title to the plot
plt.title('Distribution of Number of Genes (nFeature_RNA) by Cell Type')

# Label the x-axis and y-axis
plt.xlabel('Cell Type')
plt.ylabel('Number of Genes (nFeature_RNA)')

# Rotate the x-axis labels if they are overlapping to improve readability
plt.xticks(rotation=90)

# Adjust the plot layout to prevent elements from being cut off
plt.tight_layout()

# Display the plot
plt.show()
```

# Summary Statistics
    This will help to understand the dataset as a whole.

```{r}
#| eval: false

# shows a table of the summary statistics
print("Descriptive statistics for df_sampled numerical columns:")
display(df_sampled.describe())
```

# Progress Notes
Data subsetting
    The cohort observered so far is just the "young" (2-month-old) observations. So in order to gain the necessary insights for the research question, i.e., Young vs. Old, I need at least those two cohorts.
    
Next Steps
    The next step would be to get an appropriate "Old" cohort from the original dataset. "Streaming" the dataset allowed me to load just what I needed for the first data frame.

    May need to transform some of the visualized figures to make them more clear. Try using jitter() for the data points if they are overlapping.
        
    Include UMAP and heatmaps to help visualize both local and global structure

    Include tSNE, instead of using PCA:
    https://en.wikipedia.org/wiki/T-distributed_stochastic_neighbor_embedding

Insights from summary statistics
    The wide range observed for both `nCount_RNA` and `nFeature_RNA`, particularly the low minimum values, suggests the presence of cells with very low sequencing depth and/or gene detection. These low-quality cells might warrant further investigation or removal during quality control to avoid downstream analysis biases.
    
    To ensure reliable downstream analysis, consider implementing filtering thresholds based on these statistics (e.g., removing cells with `nCount_RNA` below a certain percentile or `nFeature_RNA` below a biologically relevant minimum) to exclude low-quality data points.


Limitations
    The original dataset is potentially too large for colab (431k rows). So I had to subset into a workable dataframe. Normally, that wouldn't be a problem if I could also subset another group to be the "Old" group for comparison, but I kept running into memory errors with colab. Iterating through tens of thousands of rows is not the way to go.

# Acknowledgments
    Dr. Sarangan (Ravi) Ravichandran and Dr. Randall Johnson from Hood College for their contributions in the form of knowledge gained through lectures, lecture notes and workbooks.
    
    Mayo clinic, background information for Glioblastoma (GBM). https://www.mayoclinic.org/diseases-conditions/glioblastoma/symptoms-causes/syc-20569077
    
    Stephanie Hicks, background information for snRNA-seq assay, "Welcome to the World of Single-Cell RNA-Sequencing". https://learn.gencore.bio.nyu.edu/single-cell-rnaseq/
    
    Gemini was used in the modification and use of some python scripts for this project.